EGGs are adhesive!\\ An algebraic account of equality saturation

### 2025-09-25 jlamp/jlamp.tex
Update jlamp.tex

\emph{Synopsis}
The paper has the following structure. 
In Section~\ref{sec:ade} we briefly recall 
the theory of $\mathcal{M}$-adhesive categories
and of kernel pairs.
In Section~\ref{sec:hyper} we present the graphical structures of our interest, 
 (labelled) hypergraphs and term graphs, and we provide a
functorial characterisation, which allows for proving their adhesivity properties.
This is expanded in Section~\ref{hypereq} for proving the $\mathcal{M}$-adhesivity
of hypergraphs 
and term graphs with equivalence and in Section~\ref{eggs} of
a few variants where equivalences are closed with respect to operator application,
thus subsuming EGGs.
%
In Section~\ref{rewriting} we put the machinery at work, showing how the optimisation steps
can be rephrased as the application of DPO rewriting rules, at the same time
highlighting the issues linked to the format of such rules.
%
Finally, in Section~\ref{conclusioni} we draw our conclusions, hint at future endeavours and offer some 
brief remarks on related works.

For the sake of readability, all the proofs appear in the appendices. 
This work is an extended version of~\cite{uovadesive}.
With respect to the conference paper, the way term graphs
are described is completely changed: they are now presented 
as an instance of the newly introduced category of 
left-monogamous hypergraphs. Also the notion of separated 
hypergraphs is original, so that Sections 3.1.1 and 3.2.1 are largely new,
and consequently so Sections 4.1.1, 4.2, and 5. 
%
Finally, also Section~6 on the use of DPO rewriting has been rewritten 
and expanded, trying to highlight the nuances of its application.
%For the sake of completeness, 
%and in order to fix the notation, we prove all the results 
%recalled in the background section, besides those that are original 
%to our work.

>>>>

\emph{Synopsis}
The paper has the following structure. 
In Section~\ref{sec:ade} we briefly recall 
the theory of $\mathcal{M}$-adhesive categories
and of kernel pairs.
In Section~\ref{sec:hyper} we present the graphical structures of our interest, 
 (labelled) hypergraphs and term graphs, and we provide a
functorial characterisation, which allows for proving their adhesivity properties.
This is expanded in Section~\ref{hypereq} for proving the $\mathcal{M}$-adhesivity
of hypergraphs and term graphs with equivalence, as well as their separated variants,
and in Section~\ref{eggs}, where equivalences are closed with respect to operator application,
thus subsuming EGGs.
%
In Section~\ref{rewriting} we put the machinery at work, showing how the optimisation steps
can be rephrased as the application of DPO rewriting rules, at the same time
highlighting the issues linked to the format of such rules.
%
Finally, in Section~\ref{conclusioni} we draw our conclusions, hint at future endeavours and offer some 
brief remarks on related works.

For the sake of readability, all the proofs appear in the appendices. 
This work is an extended version of~\cite{uovadesive}.
With respect to the conference paper, the way term graphs
are described is completely changed: they are now presented 
as an instance of the newly introduced category of 
left-monogamous hypergraphs. Also the notion of separated 
hypergraphs is original, so that Sections 3.1.1 and 3.2.1 are largely new,
and consequently so Sections 4.1.1, 4.2, and 5. 
%
Finally, also Section~6 on the use of DPO rewriting has been rewritten 
and expanded, trying to highlight the nuances concerning its application.
%For the sake of completeness, 
%and in order to fix the notation, we prove all the results 
%recalled in the background section, besides those that are original 
%to our work.
